---
title: "Introduction to twoStageLCA (Two Stage Linked Component Analysis)"
author: 
- Huan Chen
- Jinrui Liu
- Carlo Colantuoni
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to twoStageLCA (Two Stage Linked Component Analysis)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Install and load `SJD` package

To install this package in R, run the following commands:

```R
library(devtools)
install_github("CHuanSite/SJD")
```

```{r setup}
library(SJD)
```

# Single-Cell RNAseq Example

First, install the 'googleDrive' package

```{r}
install.packages('googledrive')
```

Download RNA and explaination data from Google Drive,

```{r}
library(googledrive)

## Data file 
url_data <- "https://drive.google.com/file/d/1OQovDBPwRX_O2N1GSNY8fzJn-p3-fwQV/view?usp=sharing"
drive_download(url_data, overwrite = TRUE)
unzip('data.zip')

## Explaination file
url_explaination <- 'https://drive.google.com/file/d/1S3HdygRCMvPttmVd9cix4GskWj1VPJaM/view?usp=sharing'
drive_download(url_explaination, overwrite = TRUE)
unzip('data_explaination.zip')
```

Read data into R

```{r}
library(tidyverse)

## Read in files
inVitro_bulk = read.table('1_inVitro_Bulk_Cortecon.plog2_trimNord.txt', stringsAsFactors = FALSE, header = TRUE) %>% select(-1) %>% as.matrix
inVitro_sc = read.table('2_inVitro_SingleCell_scESCdifBifurc.CelSeq_trimNord.txt', stringsAsFactors = FALSE, header = TRUE) %>% select(-1) %>% as.matrix
inVivo_bulk = read.table('3_inVivo_Bulk_BrainSpan_RNAseq_Gene_DFC_noSVA_plog2_trimNord.txt', stringsAsFactors = FALSE, header = TRUE) %>% select(-1) %>% as.matrix
inVivo_sc = read.table('4_inVivo_SingleCell_CtxDevoSC4kTopoTypoTempo_plog2_trimNord.txt', stringsAsFactors = FALSE, header = TRUE) %>% select(-1) %>% as.matrix

inVitro_bulk_scale = scale(t(scale(t(inVitro_bulk), scale = FALSE)))
inVitro_sc_scale = scale(t(scale(t(inVitro_sc), scale = FALSE)))
inVivo_bulk_scale = scale(t(scale(t(inVivo_bulk), scale = FALSE)))
inVivo_sc_scale= scale(t(scale(t(inVivo_sc), scale = FALSE)))

## legends for the 4 datasets
inVitro_bulk_exp =  read.table("1_inVitro_Bulk_Cortecon.pd.txt",stringsAsFactors = FALSE, header = T)
inVitro_sc_exp = read.table("2_inVitro_SingleCell_scESCdifBifurc.CelSeq.pd.txt", stringsAsFactors = FALSE, header = T)
inVivo_bulk_exp = read.table("3_inVivo_Bulk_BrainSpan.RNAseq.Gene.DFC.pd.txt", stringsAsFactors = FALSE, header = T)
inVivo_sc_exp = read.table("4_inVivo_SingleCell_CtxDevoSC4kTopoTypoTempo.pd.txt", stringsAsFactors = FALSE, header = T)
```

Conduct Two-stage linked component analysis

```{r}
library(SJD)
## List of datasets and group assignment and number of components
dataset = list(inVitro_bulk_scale, inVitro_sc_scale, inVivo_bulk_scale, inVivo_sc_scale)
group = list(c(1,2,3,4), c(1,2), c(3,4), c(1,3), c(2,4), c(1), c(2), c(3), c(4))
comp_num = c(2,2,2,2,2,2,2,2,2)

## Output result
twoStageLCA_res = twoStageLCA(dataset, group, comp_num)
```

Visualize the result

```{r}
par(mfrow = c(2,2))

## common component
plot(t(twoStageLCA_res$score_list[[1]][[1]]), col = inVitro_bulk_exp$color, pch = 16, xlab = "PC1", ylab = "PC2", main = "common: inVitro_bulk", cex = 1, cex.axis = 1, cex.lab = 1, cex.main = 1)
plot(t(twoStageLCA_res$score_list[[2]][[1]]), col = inVitro_sc_exp$COLORby.DCX, pch = 16, xlab = "PC1", ylab = "PC2", main = "common: inVitro_sc", cex = 1, cex.axis = 1, cex.lab = 1, cex.main = 1)
plot(t(twoStageLCA_res$score_list[[3]][[1]]), col = inVivo_bulk_exp$color, pch = 16, xlab = "PC1", ylab = "PC2", main = "common: inVivo_bulk", cex = 1, cex.axis = 1, cex.lab = 1, cex.main = 1)
plot(t(twoStageLCA_res$score_list[[4]][[1]]), col = inVivo_sc_exp$COLORby.DCX, pch = 16, xlab = "PC1", ylab = "PC2", main = "common: inVivo_sc", cex = 1, cex.axis = 1, cex.lab = 1, cex.main = 1)


par(mfrow = c(1,2))
## in vitro component
plot(t(twoStageLCA_res$score_list[[1]][[2]]), col = inVitro_bulk_exp$color, pch = 16, xlab = "PC1", ylab = "PC2", main = "common: inVitro_bulk", cex = 1, cex.axis = 1, cex.lab = 1, cex.main = 1)
plot(t(twoStageLCA_res$score_list[[2]][[2]]), col = inVitro_sc_exp$COLORby.DCX, pch = 16, xlab = "PC1", ylab = "PC2", main = "common: inVitro_sc", cex = 1, cex.axis = 1, cex.lab = 1, cex.main = 1)

## in vivo component
plot(t(twoStageLCA_res$score_list[[3]][[3]]), col = inVivo_bulk_exp$color, pch = 16, xlab = "PC1", ylab = "PC2", main = "common: inVivo_bulk", cex = 1, cex.axis = 1, cex.lab = 1, cex.main = 1)
plot(t(twoStageLCA_res$score_list[[4]][[3]]), col = inVivo_sc_exp$COLORby.DCX, pch = 16, xlab = "PC1", ylab = "PC2", main = "common: inVivo_sc", cex = 1, cex.axis = 1, cex.lab = 1, cex.main = 1)

## bulk component
plot(t(twoStageLCA_res$score_list[[1]][[4]]), col = inVitro_bulk_exp$color, pch = 16, xlab = "PC1", ylab = "PC2", main = "common: inVitro_bulk", cex = 1, cex.axis = 1, cex.lab = 1, cex.main = 1)
plot(t(twoStageLCA_res$score_list[[3]][[4]]), col = inVivo_bulk_exp$color, pch = 16, xlab = "PC1", ylab = "PC2", main = "common: inVivo_bulk", cex = 1, cex.axis = 1, cex.lab = 1, cex.main = 1)

## sc component
plot(t(twoStageLCA_res$score_list[[2]][[5]]), col = inVitro_sc_exp$COLORby.DCX, pch = 16, xlab = "PC1", ylab = "PC2", main = "common: inVitro_sc", cex = 1, cex.axis = 1, cex.lab = 1, cex.main = 1)
plot(t(twoStageLCA_res$score_list[[4]][[5]]), col = inVivo_sc_exp$COLORby.DCX, pch = 16, xlab = "PC1", ylab = "PC2", main = "common: inVivo_sc", cex = 1, cex.axis = 1, cex.lab = 1, cex.main = 1)
```

# Reference 

- Chen, H., Caffo, B., Stein-O'Brien, G., Liu, J., Langmead, B., Colantuoni, C., & Xiao, L. (2021). Two-stage Linked Component Analysis for Joint Decomposition of Multiple Biologically Related Data Sets. bioRxiv.

